---
title: "Final_DESeq2 of Mice PKD Data"
author: "Hannah Cleary"
date: "7/12/2021"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

### Section Outline

* Prep
  + Load Packages
  + Load and Convert Raw Counts
  + Load and Convert Metadata Table
  + Verify Class and Data Type
  + Insert Row Names into Data Frame
  + Initiate DESeq2 Package

## Prep

### Load Packages
```{r}
library("DESeq2")
```

### Load and Convert Raw Counts
```{r}
# read raw counts tabular data into R {fixed_raw_counts_mouse.txt}
DEpkdmice <- read.delim(file.choose(), row.names="GeneID")
# convert raw counts data frame into matrix
DEpkdmice.matrix <- as.matrix(DEpkdmice)
# remove character and double mutant columns from data matrix (col. 1-3, 10-12)
DEpkdmice.matrix <- DEpkdmice.matrix[,c(4:9)]
```

### Load and Convert Metadata Table
```{r}
# read metadata into R (data separated by commas) {SraRunTable.txt}
DEpkdmice.meta <- read.csv(file.choose(), row.names=1)
# remove double mutant rows from metadata (row. 1-6)
DEpkdmice.meta <- DEpkdmice.meta[c(1:6),]
# remove all columns from metadata except "Genotype" (col. 14)
DEpkdmice.meta <- DEpkdmice.meta["Genotype"]
DEpkdmice.meta$Genotype <- factor(DEpkdmice.meta$Genotype)
```

### Verify Class and Data Type
```{r}
# check class of DEpkdmice.meta
class(DEpkdmice.meta) # "data.frame"
# check class of DEpkdmice.matrix columns (first column)
class(DEpkdmice.matrix[,1]) # "character"
```

```{r}
# convert DEpkdmice.matrix to contain numeric columns
DEpkdmice.matrix.numeric <- apply(DEpkdmice.matrix,2,as.numeric)
# check class of DEpkdmice.matrix.numeric columns (first column)
class(DEpkdmice.matrix.numeric[,1]) # "numeric"
```

### Insert Row Names into Data Frame
```{r}
# import raw counts again without row.names argument {fixed_raw_counts_mouse.txt}
DEpkdmice.names <- read.delim(file.choose())
# save Gene IDs to a vector
GeneID <- DEpkdmice.names$GeneID
# insert row names into numeric object (DEpkdmice.matrix.numeric)
rownames(DEpkdmice.matrix.numeric) <- GeneID
```

### Initiate DESeq2 Package
```{r}
DEpkdmice.matrix.dds <- DESeqDataSetFromMatrix(countData = DEpkdmice.matrix.numeric,
                                        colData = DEpkdmice.meta,
                                        design = ~ Genotype)
# view object
DEpkdmice.matrix.dds 
```
The object created in DEpkdmice.dds is of the DESeqDataSet class with dimensions of 55385 rows and six columns. 

```{r}
# perform default analysis
DEpkdmice.matrix.dds <- DESeq(DEpkdmice.matrix.dds)
# list the coefficents
resultsNames(DEpkdmice.matrix.dds)
```

