---
title: "Final_Principal Component Analysis of Mice PKD Data"
author: "Hannah Cleary"
date: "7/9/2021"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Raw Counts Data
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149739

# Data in Published Paper
Cyclin-Dependent Kinase 1 Activity Is a Driver of Cyst Growth in Polycystic Kidney Disease
https://jasn.asnjournals.org/content/32/1/41#sec-5

# Protocol Semi-Followed
https://www.datacamp.com/community/tutorials/pca-analysis-r


## Purpose

### Section Outline

* Prep
  + Load Packages
  
## Prep

### Load Packages
```{r}
library("plyr"); library("dplyr")
# if functions are needed from both plyr and dplyr, then load plyr first, then dplyr

library("devtools")
# dependcies: ‘credentials’, ‘zip’, ‘gitcreds’, ‘ini’, ‘diffobj’, ‘gert’, ‘gh’, ‘whisker’,    
# ‘xopen’, ‘brio’, ‘praise’, ‘waldo’, ‘usethis’, ‘desc’, ‘pkgbuild’, ‘pkgload’, ‘rcmdcheck’, 
# ‘remotes’, ‘roxygen2’, ‘rversions’, ‘sessioninfo’, ‘testthat’

library("ggbiplot")
```

```{r}
# read tabular data into R {fixed_raw_counts.txt}
pkdmice <- read.delim(file.choose())
```

```{r}
# first remember the names
n <- pkdmice$GeneID

# transpose all: to make columns into rows and rows into columns
pkdmice <- as.data.frame(t(pkdmice))
colnames(pkdmice) <- n
```

```{r}
# remove first four rows of character values to generate numeric data frame
pkdmice.numeric <- pkdmice[c(5:10),]

# convert from data frame into numeric matrix 
pkdmice.matrix <- apply(as.matrix(pkdmice.numeric),2,as.numeric)

# remove columns with constant/zero column in order to rescale to unit variance
# https://stackoverflow.com/questions/40315227/how-to-solve-prcomp-default-cannot-rescale-a-constant-zero-column-to-unit-var
pkdmice.matrix.zerovariance <- pkdmice.matrix[ , which(apply(pkdmice.matrix, 2, var) != 0)]
```

```{r}
# compute the principal components
pkdmice.pca <- prcomp(pkdmice.matrix.zerovariance, center = TRUE, scale. = TRUE)
summary(pkdmice.pca)
```

```{r}
# view structure of PCA object
str(pkdmice.pca)
```
The center point ($center), scaling ($scale), standard deviation (sdev) of each principal component. The relationship (correlation or anticorrelation, etc) between the initial variables and the principal components ($rotation). The values of each sample in terms of the principal components ($x).

```{r}
# plot without arrow lines
ggbiplot(pkdmice.pca, var.axes = FALSE)

# plot without arrow lines and application of row names (nine samples)
ggbiplot(pkdmice.pca, var.axes = FALSE, labels=rownames(pkdmice.numeric))
```
now you can see which samples are similar to one another

```{r}
# put samples into categories based on presence/absence of mutation in pkdmice.numeric, first three rows are control, second three rows are PC2 mutants, and last two rows are double mutants
pkdmice.genetype <- c(rep("Wildtype",3), rep("Pkd2 Knockout",3))

# ellipse = TRUE draws an ellipse around each group
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype) + xlab("Standardized PC1 (30.7% Explained Variance)") + ylab("Standardized PC2 (21.7% Explained Variance)") + ggtitle("Principal Component Analysis for \n  Wildtype and Pkd2 Knockout Mice Data") + theme(plot.title=element_text(face="bold", hjust=0.5, size=15), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"), axis.title=element_text(size=10), legend.position="right") + labs(color="Sample Groups") + expand_limits(x = c(-2:2))
```

```{r}
# save plot as jpeg
jpeg('MousePKDPCA_PC12_NoDBL.jpg')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)
dev.off()

# save plot as png
png('MousePKDPCA_PC12_NoDBL.png')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)
dev.off()

# save plot as pdf
pdf('MousePKDPCA_PC12_NoDBL.pdf')
pdf('MousePKDPCA_Final')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype) + xlab("Standardized PC1 (30.7% Explained Variance)") + ylab("Standardized PC2 (21.7% Explained Variance)") + ggtitle("Principal Component Analysis for \n  Wildtype and Pkd2 Knockout Mice Data") + theme(plot.title=element_text(face="bold", hjust=0.5, size=15), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"), axis.title=element_text(size=10), legend.position="right") + labs(color="Sample Groups") + expand_limits(x = c(-2:2))
dev.off()
```

```{r}
ggbiplot(pkdmice.pca,ellipse=TRUE, choices = 3:4, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)

# save plot as jpeg
jpeg('MousePKDPCA_PC34_NoDBL.jpg')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)
dev.off()

# save plot as png
png('MousePKDPCA_PC34_NoDBL.png')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)
dev.off()

# save plot as pdf
pdf('MousePKDPCA_PC34_NoDBL.pdf')
ggbiplot(pkdmice.pca,ellipse=TRUE, var.axes = FALSE, labels=rownames(pkdmice.numeric), groups=pkdmice.genetype)
dev.off()
```
